What is an Ancestral Recombination Graph?

### 2022-02-13 paper.tex
Mention "big" vs "little" ARG earlier

% NOTE: it would probably be helpful to have a drawing of the ARG process
% in-progress here to help us talk about it. What is the full state
% that is remembered as we go through the process? It does help to think
% about the process as choosing edges to merge, not nodes.
The Ancestral Recombination Graph (ARG) was later introduced by
Griffiths~\citep{griffiths1991two,ethier1990two,griffiths1996ancestral,
griffiths1997ancestral}, and is a
closely related and complementary treatment of the same process. Where
Hudson's algorithm is focused on the efficient \emph{simulation} of the
process, the original ARG literature is focused on mathematical
results about the stochastic process. Griffiths and colleagues formulated the
ARG as a branching-coalescing stochastic process. We imagine a graph in which
each edge corresponds to an extant lineage and nodes are events in the
process (the initial $n$ leaf nodes are ``sampling'' events). As in Hudson's
algorithm, common ancestor events occur at rate $\binom{k}{2}$ when there
are $k$ lineages present. We choose two lineages (edges) uniformly, and merge them
into a common ancestor lineage. Recombination events are different
to Hudson's algorithm, however. Rather than occuring at a rate which depends on
the specific distribution of ancestral material among lineages, the
rate of recombination is simply $k \rho (m - 1)$ in the original ARG
formulation. At a recombination event we choose a lineage (edge) uniformly, and a
breakpoint $0 < x < L$ uniformly on its genome. We terminate the edge at a
node, record the breakpoint and start two new edges from this node. The process
then continues until there is only one lineage left (the Grand Most Recent
Common Ancestor, GMRCA), which is guaranteed to
happen in finite time because of the linear vs quadratic rates of branching
and coalescing. The graph structure and breakpoints associated with
recombination nodes provides sufficient information to later recover the marginal
trees (see the next section for more details and discussion on this point).

The state-space of this process is much simpler than Hudson's algorithm, which
greatly facilitates mathematical reasoning. This simplicity comes at a
substantial cost, however, if we wish to use it as a practical means of
simulating recombinant ancestry. The ARG is vast, and any realisation
for even moderate levels of recombination is far too large to be of practical
use. The number of events in the ARG back to the GMRCA
is $O(e^\rho)$~\citep{griffiths1997ancestral}, whereas the number
of events required to simulate the complete ancestry of a sample
in Hudson's algorithm is
$O(\rho^2)$~\citep{hein2004gene,baumdicker2021efficient}.
This disparity in the number of events in the two formulations is
because the majority of the events that occur in the ARG do
not affect the genetic ancestry of the sample in any way. Recombination
events that occur outside of ancestral material do not have any bearing
on the ancestry of the sample, and so the structure is hugely redundant.
As~\cite{wiuf1999recombination} note,
``an `ancestral' sequence in the birth and death process
need not have any genetic material in common with a
sequence descended from it.''

@@

This original formulation which does not keep track of the distribution
of ancestral material among lineages is the so-called ``Big'' ARG.
The ``little'' ARG is the subset of the Big ARG traversed by
Hudson's algorithm (\citet{wiuf1999recombination} refer to this as HUD).
[Review of other work done on the ARG stochastic process here.
Discuss \citet{wiuf1999ancestry}. Wiuf and Hein's classical pair of papers
clarified the relationship between Hudson's algorithm and the ARG.
Explain that ARG simulation is inherently less efficient than Hudson's
algorithm because we must track all lineages, even those that correspond
to sections that have coalesced, or we won't have a complete ARG.
This might be tricky to communicated without a figure, but is an
esoteric point perhaps not worth spending a figure on.]

@@

The idea of using branching-coalescing stochastic processes
to model genetic ancestral trees has also found application in natural selection.
The Ancestral Selection Graph (ASG)~\citep{krone1997ancestral,neuhauser1997genealogy}
uses dynamics identical to the ``big" ARG to simulate an ensemble of correlated
potential ancestral trees. Weak genic selection is incorporated by sampling a
true ancestry from the ensemble in a non-uniform way.
There is no ASG analogue of the more computationally tractable ``little" ARG,
though some gains in tractability can be made by considering typed
lineages~\citep{etheridge2009coalescent} or by leveraging perfect simulation
techniques when recurrent mutation is present\citep{fearnhead2001perfect}.
Extensions of the ASG have been developed to frequency-dependent
selection~\citep{neuhauser1999ancestral, gonzalezcasanova2018duality},
unlinked chromosomes~\citep{fearnhead2003ancestral}, recombining
loci whereupon branching is due to both selection and
recombination~\citep{donnelly1999genealogical}, and high fecundity
reproduction~\citep{gonzalezcasanova2018duality, koskela2019robust}.

@@

The term \emph{phylogenetic network} has been broadly accepted to describe the most general
type of genealogical graph, defined by \citet{huson2010phylogenetic} as ``any graph used to
represent evolutionary relationships (either abstractly or explicitly) between a set of taxa
that labels some of its nodes" (although more narrow and context-dependent definitions can also
be found in the literature). Explicit rooted phylogenetic networks are rooted DAGs,
with leaves labelled by the taxa (species, groups, individuals or sequences), and nodes classed
as either reticulation nodes (with in-degree $\geq 2$) or tree nodes.

@@

\subsection*{Model-based methods}
An alternative approach is to treat the ancestry as a latent parameter to be averaged out
by Monte Carlo methods, based either on importance sampling
\citep{griffiths1996ancestral, fearnhead2001estimating, jenkins2011inference}
or MCMC \citep{kuhner2000maximum, nielsen2000estimation, wang2008bayesian, fallon2013acg}.
These methods invariably operated on a representation of the ``little ARG", typically
generated until the grand MRCA. They are extremely computationally expensive,
and applicable to at most hundreds of samples consisting of tens of kilobases with
human-like parameters.

@@

A very recent MCMC method called \texttt{ARGInfer} has resulted in the first
improvement in the scalability of Monte Carlo methods for the exact ARG model
by making use of the tree sequence representation of marginal trees, suitably enriched
to facilitate the computations needed for MCMC sampling \citep{mahmoudi2021inference}.
While not competitive with the scalability of \texttt{ARGWeaver}, \texttt{ARGInfer} extends
the feasible range of sequence lengths by at least an order of magnitude when
compared with methods which sample realisations of the exact ``little ARG".

>>>>

% NOTE: it would probably be helpful to have a drawing of the ARG process
% in-progress here to help us talk about it. What is the full state
% that is remembered as we go through the process? It does help to think
% about the process as choosing edges to merge, not nodes.
The Ancestral Recombination Graph (ARG) was later introduced by
Griffiths~\citep{griffiths1991two,ethier1990two,griffiths1996ancestral,
griffiths1997ancestral}, and is a
closely related and complementary treatment of the same process. Where
Hudson's algorithm is focused on the efficient \emph{simulation} of the
process, the original ARG literature is focused on mathematical
results about the stochastic process. Griffiths and colleagues used the word
``ARG'' to refer to both a specific branching-coalescing stochastic process and
the graph structure derived from this process. For clarity, we will refer to the
process as the Griffiths process, or the coalescent with recombination (CwR),
and following XX-Who?-XX, we refer to the the structure as the ``big ARG'', as
opposed to the ``little ARG'' traversed by Hudson's algorithm
(\citet{wiuf1999recombination} refer to these structures as ARG and HUD respectively).

In the Griffiths formulation, each edge in the graph corresponds to an extant
lineage and nodes are events in the process (the initial $n$ leaf nodes are
``sampling'' events). As in Hudson's
algorithm, common ancestor events occur at rate $\binom{k}{2}$ when there
are $k$ lineages present. We choose two lineages (edges) uniformly, and merge them
into a common ancestor lineage. Recombination events are different
to Hudson's algorithm, however. Rather than occuring at a rate which depends on
the specific distribution of ancestral material among lineages, the
rate of recombination is now simply $k \rho (m - 1)$. At a recombination event
we choose a lineage (edge) uniformly, and a
breakpoint $0 < x < L$ uniformly on its genome. We terminate the edge at a
node, record the breakpoint and start two new edges from this node. The process
then continues until there is only one lineage left (the Grand Most Recent
Common Ancestor, GMRCA), which is guaranteed to
happen in finite time because of the linear vs quadratic rates of branching
and coalescing. The graph structure and breakpoints associated with
recombination nodes provides sufficient information to later recover the marginal
trees (see the next section for more details and discussion on this point).

The state-space of the Griffiths process is much simpler than Hudson's algorithm,
which greatly facilitates mathematical reasoning. This simplicity comes at a
substantial cost, however, if we wish to use it as a practical means of
simulating recombinant ancestry. The big ARG is vast, and any realisation
for even moderate levels of recombination is far too large to be of practical
use. The number of events in the big ARG, all the way back to the GMRCA
is $O(e^\rho)$~\citep{griffiths1997ancestral}, whereas the number
of events required to simulate the little ARG is
$O(\rho^2)$~\citep{hein2004gene,baumdicker2021efficient}.
This disparity in the number of events in the two formulations is
because the majority of the events that occur in the big ARG do
not affect the genetic ancestry of the sample in any way. Recombination
events that occur outside of ancestral material do not have any bearing
on the ancestry of the sample, and so the structure is hugely redundant.
As~\cite{wiuf1999recombination} note,
``an `ancestral' sequence in the birth and death process
need not have any genetic material in common with a
sequence descended from it.''

@@

[Review of other work done on the ARG stochastic process here.
Discuss \citet{wiuf1999ancestry}. Wiuf and Hein's classical pair of papers
clarified the relationship between Hudson's algorithm and the ARG.
Explain that ARG simulation is inherently less efficient than Hudson's
algorithm because we must track all lineages, even those that correspond
to sections that have coalesced, or we won't have a complete ARG.
This might be tricky to communicated without a figure, but is an
esoteric point perhaps not worth spending a figure on.]

@@

The idea of using branching-coalescing stochastic processes
to model genetic ancestral trees has also found application in natural selection.
The Ancestral Selection Graph (ASG)~\citep{krone1997ancestral,neuhauser1997genealogy}
uses dynamics identical to the ``big'' ARG to simulate an ensemble of correlated
potential ancestral trees. Weak genic selection is incorporated by sampling a
true ancestry from the ensemble in a non-uniform way.
There is no ASG analogue of the more computationally tractable ``little'' ARG,
though some gains in tractability can be made by considering typed
lineages~\citep{etheridge2009coalescent} or by leveraging perfect simulation
techniques when recurrent mutation is present\citep{fearnhead2001perfect}.
Extensions of the ASG have been developed to frequency-dependent
selection~\citep{neuhauser1999ancestral, gonzalezcasanova2018duality},
unlinked chromosomes~\citep{fearnhead2003ancestral}, recombining
loci whereupon branching is due to both selection and
recombination~\citep{donnelly1999genealogical}, and high fecundity
reproduction~\citep{gonzalezcasanova2018duality, koskela2019robust}.

@@

The term \emph{phylogenetic network} has been broadly accepted to describe the most general
type of genealogical graph, defined by \citet{huson2010phylogenetic} as ``any graph used to
represent evolutionary relationships (either abstractly or explicitly) between a set of taxa
that labels some of its nodes'' (although more narrow and context-dependent definitions can also
be found in the literature). Explicit rooted phylogenetic networks are rooted DAGs,
with leaves labelled by the taxa (species, groups, individuals or sequences), and nodes classed
as either reticulation nodes (with in-degree $\geq 2$) or tree nodes.

@@

\subsection*{Model-based methods}
An alternative approach is to treat the ancestry as a latent parameter to be averaged out
by Monte Carlo methods, based either on importance sampling
\citep{griffiths1996ancestral, fearnhead2001estimating, jenkins2011inference}
or MCMC \citep{kuhner2000maximum, nielsen2000estimation, wang2008bayesian, fallon2013acg}.
These methods invariably operated on a representation of the ``little ARG'', typically
generated until the grand MRCA. They are extremely computationally expensive,
and applicable to at most hundreds of samples consisting of tens of kilobases with
human-like parameters.

@@

A very recent MCMC method called \texttt{ARGInfer} has resulted in the first
improvement in the scalability of Monte Carlo methods for the exact ARG model
by making use of the tree sequence representation of marginal trees, suitably enriched
to facilitate the computations needed for MCMC sampling \citep{mahmoudi2021inference}.
While not competitive with the scalability of \texttt{ARGWeaver}, \texttt{ARGInfer} extends
the feasible range of sequence lengths by at least an order of magnitude when
compared with methods which sample realisations of the exact ``little ARG''.
